---
title: "ROV Survey Data Compilation — Yelloweye Rockfish Stock Assessment"
author: "Laura Coleman (laura.coleman@alaska.gov)"
date: "Last modified: July 16, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This document compiles all of the ROV survey data. It includes raw species
review data for ROV surveys 2012–2023.

The goal here is to create one file with **ALL** of the raw ROV line transect
data for the yelloweye rockfish stock assessment survey. This includes all
transects, with all species. Kelli or Phil compiled a lot of this data
already, so this starts with the outputs they created and recombines raw
files where needed.

> **Note:** This code should not be used to get the number of yelloweye
> rockfish (YE) for the stock assessment survey. Yelloweye rockfish from the
> "bad" segments are believed to be removed and not counted.
> **FLAG!** This is something that still needs follow-up.

## ROV Surveys for DSR Stock Assessment

| Year | Area |
|------|------|
| 2012 | CSEO |
| 2013 | SSEO |
| 2014 | CANCELED |
| 2015 | EYKT |
| 2016 | CSEO, NSEO |
| 2017 | EYKT |
| 2018 | NSEO, CSEO, SSEO |
| 2019 | EYKT |
| 2020 | SSEO |
| 2021 | CANCELLED |
| 2022 | CSEO, NSEO |
| 2023 | EYKT |

The ROV program was suspended after 2023 due to the retirement of ROV pilot
Mike Byerly.

## How This Document Is Organized

Each survey year below is broken into four parts:

- **Raw Data** — importing the original species review file(s) for that year
- **Data Issues** — specific corrections applied in code (wrong dive/transect
  numbers, missing columns, mismatched filenames, etc.)
- **Known Problems** — unresolved flags and open questions about the data
  that have **not** been fixed in code, and may need follow-up
- **Data Standardization** — aligning column names/types to the common
  template (`EYKT_2023`) so all years can eventually be combined

After all years are processed, the document moves on to combining,
cleaning, exploring, and analyzing the full dataset.

# Setup

```{r source-helper}
source('scripts/helper.r')

###  set plotting theme to use TNR  ###
# #font_import() #remove # to run this but only do this one time - it takes a while
# loadfonts(device="win")
# windowsFonts(Times=windowsFont("TT Times New Roman"))
# theme_set(theme_bw(base_size=18,base_family='Times New Roman')
#           +theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()))
```

# 2012 — CSEO

## Raw Data

`Filepath: C:\Users\lscoleman\Documents\R code - Working Files\seak_dsr_survey\data\SPECIES_REVIEW_DATA\2012_CSEO\FishVideoReview_Feb2_2013ROV lengths_kg.csv`

```{r import-2012}
CSEO_2012_ALLdata <- read_csv("data/SPECIES_REVIEW_DATA/2012_CSEO/species_CSEO_2012.csv")
```

The 2012 ROV Distance Analysis surveyed 46 transects and observed a total of 118 yelloweye rockfish, consisting of 112 adults and 6 subadults greater than 350 mm. The encounter rate was 0.003 yelloweye per unit transect length, down from 0.005 in 2007, with an average of 2.6 fish observed per transect compared to 5 in 2007. The resulting density estimate was 752 yelloweye per km², a decline from the 1,068 yelloweye per km² estimated in 2007, with a coefficient of variation of 0.125 indicating high precision in the estimate. The final model selected was a hazard-rate cosine key function with no binning and no truncation, chosen because it produced the lowest AIC, the strongest goodness-of-fit results (via K-S and chi-square tests), and the best fit on the Q-Q plot compared to alternative models.
The interpretation of these results is that the decline in density from 2007 to 2012 likely reflects a real drop in the yelloweye rockfish population rather than a methodological artifact. This decline is attributed to five years of fishing pressure on the population, including sport fishing (both charter and unguided), bycatch in the commercial halibut fishery, and a directed DSR fishery that removed roughly 86,000 pounds of yelloweye in 2012. Because yelloweye rockfish are slow-growing, late-maturing, and exhibit high site fidelity, recruitment and repopulation—whether from within the population or from other management areas—would be expected to occur slowly, reinforcing the conclusion that this represents a genuine, lasting population decline. Additionally, the analysis validated the ROV as a precise and reliable method for estimating yelloweye density, supporting its continued use as a substitute for submersible-based surveys.
## Data Issues

None identified in code for this year.

## Known Problems

**FLAG!** I cannot find the original files from EventMeasure, so I am unable
to combine everything myself from scratch — this file is being taken as-is.

## Data Standardization

`EYKT_2023` is used as the running comparison template throughout the
standardization sections in this document.

```{r standardize-2012}
cols_cseo_2012 <- names(CSEO_2012)
cols_eykt_2023 <- names(EYKT_2023)

# Get max length between the two
max_len <- max(length(cols_cseo_2012), length(cols_eykt_2023))

# Make lists to the same length
cols_cseo_2012 <- c(cols_cseo_2012, rep(NA, max_len - length(cols_cseo_2012)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols1 <- data.frame(
  in_CSEO_2012 = cols_cseo_2012,
  in_EYKT_2023 = cols_eykt_2023);compare_cols1


CSEO_2012 <- CSEO_2012 %>% 
  mutate(mgmt_area = "CSEO",
         dive_type = "Line",
         time_hms = NA,
         period_time_hms = NA,
         event_time_hh_mm_ss = NA,
         time_akdt = NA) %>% #time_akdt is in 2015 EYKT so adding it here 
  rename(precision_mm = precision,
         rms_mm = rms_1_mm,
         op_code = opcode,
         depth = depth_not_used,
         comment = trip_comment,
         dive = dive_no,
         transect_number = transect_no,
         number = specimen_number,
         comment_1 = species_comment)

EYKT_2023 <- EYKT_2023 %>% 
  mutate(mgmt_area = "EYKT",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         time_akdt = NA) #time_akdt is in 2015 EYKT so adding it here )
```

# 2013 — SSEO

## Raw Data

`Filepath: M:\ROVSurvey\2013\species_SSEO_2013`

This was created in 2021, so I am assuming it was created by Phil or Kelli
but there are 8 dives missing according to the dive log.

The 2013 ROV Distance Analysis notes have been saved in the documents folder.

```{r import-2013}
SSEO_2013_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2013_SSEO/species_SSEO_2013.csv") %>% 
  mutate(Time..mins. = NA,
         Period.time..mins. = NA) #adding these columns to match the missing data

SSEO_2013_missing_data <- read.csv("data/SPECIES_REVIEW_DATA/2013_SSEO/species_SSEO_2013_missing_dives.csv") %>% 
  rename(Comment=Comment...20,
         Comment.1=Comment...37)

SSEO_2013_ALLdata <- rbind(SSEO_2013_ALLdata,SSEO_2013_missing_data)
```

## Data Issues

- Added `Time..mins.` and `Period.time..mins.` as `NA` to the main file so
  its columns match the missing-dives file before combining.
- Renamed mismatched comment columns (`Comment...20`, `Comment...37`) in the
  missing-dives file to `Comment` and `Comment.1` so they align.

## Known Problems

- **FLAG!** I cannot find the original files from EventMeasure, so I am
  unable to combine everything myself from scratch.
- **FLAG!** I found EventMeasure files for the missing dives but did not find
  documentation on why these dives were not included in the
  `species_SSEO_2013` file originally. I want to revisit this and confirm
  those dives were included in the actual 2013 assessment.

## Data Standardization

```{r standardize-2013}
cols_sseo_2013 <- names(SSEO_2013)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_sseo_2013), length(cols_eykt_2023))

# Make lists to the same length
cols_sseo_2013 <- c(cols_sseo_2013, rep(NA, max_len - length(cols_sseo_2013)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols2 <- data.frame(
  in_SSEO_2013 = cols_sseo_2013,
  in_EYKT_2023 = cols_eykt_2023);compare_cols2

SSEO_2013 <- SSEO_2013 %>% 
  mutate(mgmt_area = "SSEO",
         dive_type = "Line",
         rms_2_mm = NA,
         time_akdt = NA) %>% #time_akdt is in 2015 EYKT so adding it here
  rename(rms_mm = rms_1_mm,
         dive = dive_no,
         transect_number = transect_no,
         event_time_hh_mm_ss = event_time)
```

# 2015 — EYKT

## Raw Data

I combined the raw files for this survey — see code:
`2015_EYKT_MergingCSVFilesCode_SpeciesReviewData`. The 2015 ROV Distance
Analysis notes have been saved in the documents folder.

```{r import-2015}
EYKT_2015_ALLdata <- read.csv("outputs/species_review/SPECIES_EYKT_2015.csv") 
```

## Data Issues

- Filtered out rows with `NA` for `year` during standardization (see below).

## Known Problems

None identified for this year.

## Data Standardization

```{r standardize-2015}
cols_eykt_2015 <- names(EYKT_2015)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_eykt_2015), length(cols_eykt_2023))

# Make lists to the same length
cols_eykt_2015 <- c(cols_eykt_2015, rep(NA, max_len - length(cols_eykt_2015)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols3 <- data.frame(
  in_EYKT_2015 = cols_eykt_2015,
  in_EYKT_2023 = cols_eykt_2023);compare_cols3

EYKT_2015 <- EYKT_2015 %>% 
  mutate(mgmt_area = "EYKT",
         dive_type = "Line",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA) %>%  
  rename(dive = dive_no,
         transect_number = transect_no) %>% 
  filter(!is.na(year))
```

# 2016 — CSEO & NSEO

`File path: M:\ROVSurvey\2016`

The 2016 ROV Distance Analysis notes have been saved in the documents folder.

## Raw Data

### CSEO

```{r import-2016-cseo}
CSEO_2016_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2016_CSEO/species_CSEO_2016.csv")
```

### NSEO

```{r import-2016-nseo}
NSEO_2016_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2016_NSEO/species_NSEO_2016.csv")
```

## Data Issues

- **NSEO:** filtered out a rosethorn rockfish record with no location data
  (see standardization step below).

## Known Problems

None identified for this year.

## Data Standardization

### CSEO

```{r standardize-2016-cseo}
cols_cseo_2016 <- names(CSEO_2016)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_cseo_2016), length(cols_eykt_2023))

# Make lists to the same length
cols_cseo_2016 <- c(cols_cseo_2016, rep(NA, max_len - length(cols_cseo_2016)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols4 <- data.frame(
  in_CSEO_2016 = cols_cseo_2016,
  in_EYKT_2023 = cols_eykt_2023);compare_cols4

CSEO_2016 <- CSEO_2016 %>% 
  mutate(mgmt_area = "CSEO",
         dive_type = "Line",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA) %>%  
  rename(dive = dive_no,
         transect_number = transect_no)
```

### NSEO

```{r standardize-2016-nseo}
cols_nseo_2016 <- names(NSEO_2016)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_nseo_2016), length(cols_eykt_2023))

# Make lists to the same length
cols_nseo_2016 <- c(cols_nseo_2016, rep(NA, max_len - length(cols_nseo_2016)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols5 <- data.frame(
  in_NSEO_2016 = cols_nseo_2016,
  in_EYKT_2023 = cols_eykt_2023);compare_cols5

NSEO_2016 <- NSEO_2016 %>% 
  mutate(mgmt_area = "NSEO",
         dive_type = "Line",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         event_time_hh_mm_ss = NA) %>%  
  rename(dive = dive_no,
         transect_number = transect_no) %>% 
  filter(!is.na(year)) #there is a rosethorn rf with no location data
```

# 2017 — EYKT

## Raw Data

`File path: M:\ROVSurvey\2017`

The 2017 ROV Distance Analysis notes have been saved in the documents folder.

```{r import-2017}
EYKT_2017_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2017_EYKT/species_EYKT_2017.csv")
```

## Data Issues

- Dive 16 (`SL_2017EYKT_Dive_16_11-05-02.000.avi`) was reclassified as
  `dive_type = "Exploratory"` (see standardization step below) — exploratory
  dives are later excluded from the final cleaned dataset.

## Known Problems

None identified for this year.

## Data Standardization

```{r standardize-2017}
cols_eykt_2017 <- names(EYKT_2017)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_eykt_2017), length(cols_eykt_2023))

# Make lists to the same length
cols_eykt_2017 <- c(cols_eykt_2017, rep(NA, max_len - length(cols_eykt_2017)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols6 <- data.frame(
  in_EYKT_2017 = cols_eykt_2017,
  in_EYKT_2023 = cols_eykt_2023);compare_cols6

EYKT_2017 <- EYKT_2017 %>% 
  mutate(mgmt_area = "EYKT",
         dive_type = "Line",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         event_time_hh_mm_ss = NA,
         time_akdt =  NA) %>%  
  rename(dive = dive_no,
         transect_number = transect_no) %>% 
  mutate(dive_type = case_when(
    filename == "SL_2017EYKT_Dive_16_11-05-02.000.avi" ~ "Exploratory", TRUE ~ as.character(dive_type)))
```

# 2018 — CSEO, NSEO, SSEO

- `File path: M:\ROVSurvey\2018\CSEO\R Data Files\2018_cseo_species`
- `File path: M:\ROVSurvey\2018\NSEO\NSEO_2018_species`
- `File path: M:\ROVSurvey\2018\SSEO\Species Review Files\2018ROV Survey_FishVideoReview_SSEO`

It looks like dives 25 and 26 are missing, but these dives were BRF
experimental dives. There are `.text` files in the
`ROVSurvey/2018/CSEO/Stereo Files` folder for those experimental dives.

## Raw Data

### CSEO

```{r import-2018-cseo}
CSEO_2018_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2018_CSEO/species_CSEO_2018.csv")
```

### NSEO

```{r import-2018-nseo}
#columns for NSEO 
cols <- c("Filename","Frame","Time..HMS.","Period","Period.time..HMS.","Length..mm.","Precision..mm.","RMS..mm.",
          "Range..mm.","Direction..deg.","Horz..Dir...deg.","Vert..Dir...deg.","Mid.X..mm.","Mid.Y..mm.","Mid.Z..mm.",
          "OpCode","TapeReader","Depth","Comment","Year","Dive","Transect.Number","Dive.Type","Family","Genus",
          "Species","Code","Number","Stage","Activity","Check.ID","Comment.1","Event.Time.HH.MM.SS")

NSEO_2018_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2018_NSEO/species_NSEO_2018.csv") %>% 
  mutate(Period = NA, 
         Period.time..HMS. = NA,
         Comment.1 = NA,
         Event.Time.HH.MM.SS = NA) %>% 
  select(any_of(cols))

NSEO_2018_dive26 <- read.csv("data/SPECIES_REVIEW_DATA/2018_NSEO/NSEO_2018_Dive26.csv") %>% 
  select(! c(X,X.1,X.2,X.3)) %>% 
  select(any_of(cols))

NSEO_2018_ALLdata <- rbind(NSEO_2018_ALLdata,NSEO_2018_dive26)

unique(NSEO_2018_ALLdata$Dive)

#Export NSEO 2018 to check time codes
write.csv(NSEO_2018_ALLdata,"outputs/nseo_2018_alldata.csv")
```

### SSEO

```{r import-2018-sseo}
SSEO_2018_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2018_SSEO/species_SSEO_2018.csv") %>% 
  # need to fix the dive and transect numbers for dive 16
  mutate(DIVE_NO = case_when(
    FILENAME == "SL_2018SSEO_Dive_16_19-03-14.000.avi" ~ 16, TRUE ~ DIVE_NO)) %>%  
  mutate(TRANSECT_NUMBER = case_when(
    FILENAME == "SL_2018SSEO_Dive_16_19-03-14.000.avi" ~ 64,TRUE ~ TRANSECT_NUMBER)) %>%
  # need to fix the dive and transect numbers for dive 21
  mutate(DIVE_NO = case_when(
    FILENAME == "SL_2018SSEO_Dive_21_10-59-57.000.avi" ~ 21,TRUE ~ DIVE_NO)) %>%  
  mutate(TRANSECT_NUMBER = case_when(
    FILENAME == "SL_2018SSEO_Dive_21_10-59-57.000.avi" ~ 69,TRUE ~ TRANSECT_NUMBER))

SSEO_2018_ALLdata %>% summarise(distinct_dives = n_distinct(DIVE_NO))
```

## Data Issues

- **NSEO:** dive 26 was found in a separate file (`NSEO_2018_Dive26.csv`) and
  appended to the main NSEO dataset, after dropping its junk `X`/`X.1`/`X.2`/`X.3`
  columns and aligning it to the same column set.
- **SSEO:** dive and transect numbers were corrected for two mislabeled
  filenames — dive 16 (transect 64) and dive 21 (transect 69).

## Known Problems

- **FLAG!!!** NSEO dive 26 is missing from the original spreadsheet, and I
  don't think it was included in the assessment — it's not in any of the
  combined species files for 2018 NSEO. (It has since been appended above,
  but the reason it was originally excluded is still unclear.)
- **SSEO:** the transect numbers recorded on the dive log from the field do
  not match what was recorded in the species review files — I am assuming
  the species review files are correct. May want to cross-reference this
  with the GPS locations.
- **SSEO:** dive 8 is missing because there was a note that the data was bad
  and to not use it, but there was no other explanation on why the data was
  bad (that I could find anyways!!). **FLAG!** Do we want to include dive 8
  for completeness? I could not figure out why the data was bad, so I am
  worried about adding it and it accidentally being used for analysis.
- **SSEO:** there are two other dives (34 and 35) on the dive log that are
  not in the species review. These dives were never analyzed in
  EventMeasure and do not belong to another area.
- With all this in mind, 32 transects were used in the SEO DSR stock
  assessment but 35 were attempted.

## Data Standardization

### NSEO

```{r standardize-2018-nseo}
cols_nseo_2018 <- names(NSEO_2018)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_nseo_2018), length(cols_eykt_2023))

# Make lists to the same length
cols_nseo_2018 <- c(cols_nseo_2018, rep(NA, max_len - length(cols_nseo_2018)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols7 <- data.frame(
  in_NSEO_2018 = cols_nseo_2018,
  in_EYKT_2023 = cols_eykt_2023);compare_cols7

NSEO_2018 <- NSEO_2018 %>% 
  mutate(mgmt_area = "NSEO",
         dive_type = "Line",
         time_mins = NA,
         period = NA,
         period_time_mins = NA,
         period_time_hms = NA,
         rms_2_mm = NA,
         time_akdt =  NA,
         event_time_hh_mm_ss = NA,
         comment_1 = NA)
```

### CSEO

```{r standardize-2018-cseo}
cols_cseo_2018 <- names(CSEO_2018)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_cseo_2018), length(cols_eykt_2023))

# Make lists to the same length
cols_cseo_2018 <- c(cols_cseo_2018, rep(NA, max_len - length(cols_cseo_2018)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols8 <- data.frame(
  in_CSEO_2018 = cols_cseo_2018,
  in_EYKT_2023 = cols_eykt_2023);compare_cols8

CSEO_2018 <- CSEO_2018 %>% 
  select(!dive) %>% #"dive" is "management area"_ "dive number", which is not helpful here
  mutate(mgmt_area = "CSEO",
         dive_type = "Line",
         time_mins = NA,
         period = NA,
         period_time_mins = NA,
         period_time_hms = NA,
         rms_2_mm = NA,
         time_akdt =  NA,
         event_time_hh_mm_ss = NA) %>%  
  rename(op_code = opcode,
         tape_reader = tapereader,
         dive = dive_no)
```

### SSEO

```{r standardize-2018-sseo}
cols_sseo_2018 <- names(SSEO_2018)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_sseo_2018), length(cols_eykt_2023))

# Make lists to the same length
cols_sseo_2018 <- c(cols_sseo_2018, rep(NA, max_len - length(cols_sseo_2018)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols9 <- data.frame(
  in_SSEO_2018 = cols_sseo_2018,
  in_EYKT_2023 = cols_eykt_2023);compare_cols9

SSEO_2018 <- SSEO_2018 %>% 
  select(!dive) %>% #"dive" is "management area"_ "dive number", which is not helpful here
  mutate(mgmt_area = "SSEO",
         dive_type = "Line",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         time_akdt =  NA) %>%
  rename(op_code = opcode,
         tape_reader = tapereader,
         dive = dive_no)
```

# 2019 — EYKT

## Raw Data

`File path: M:\ROVSurvey\2019\EYKT\R Code\EYKT_2019\EYKT_2019_species`

There is not a document for the analysis beyond the SAFE report.

```{r import-2019}
EYKT_2019_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2019_EYKT/species_EYKT_2019.csv") %>% 
  mutate(Dive = case_when(
  Filename == "SL_2019_EYKT_Dive_12_09-27-40.000.avi" ~ 12, TRUE ~ Dive))
```

## Data Issues

- Dive number was corrected to 12 for filename
  `SL_2019_EYKT_Dive_12_09-27-40.000.avi`.

## Known Problems

- **FLAG!** There is no analysis documentation for this year beyond the SAFE
  report.

## Data Standardization

```{r standardize-2019}
cols_eykt_2019 <- names(EYKT_2019)
cols_eykt_2023 <- names(EYKT_2023) #should include the changes from the above step

# Get max length between the two
max_len <- max(length(cols_eykt_2019), length(cols_eykt_2023))

# Make lists to the same length
cols_eykt_2019 <- c(cols_eykt_2019, rep(NA, max_len - length(cols_eykt_2019)))
cols_eykt_2023 <- c(cols_eykt_2023, rep(NA, max_len - length(cols_eykt_2023)))

# Combine into a data frame
compare_cols10 <- data.frame(
  in_EYKT_2019 = cols_eykt_2019,
  in_EYKT_2023 = cols_eykt_2023);compare_cols10

EYKT_2019 <- EYKT_2019 %>% 
  mutate(mgmt_area = "EYKT",
         time_mins = NA,
         period = NA,
         period_time_mins = NA,
         period_time_hms = NA,
         rms_2_mm = NA,
         time_akdt =  NA)
```

# 2020 — SSEO

## Raw Data

`File path: M:\ROVSurvey\2020\R Code\Data\SSEO_2020_Species`

The 2020 ROV Distance Analysis notes have been saved in the documents folder.

```{r import-2020}
SSEO_2020_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2020_SSEO/species_SSEO_2020.csv") %>%
  #the transect number and dive number were flipped for dive 31
  mutate(Dive = case_when(
    Filename == "SL_2020_SSEO_Dive_31_09-34-49.000.avi" ~ 31, TRUE ~ Dive),
    Transect.Number = case_when(
      Filename == "SL_2020_SSEO_Dive_31_09-34-49.000.avi" ~ 18, TRUE ~ Transect.Number))

unique(SSEO_2020_ALLdata$Dive)
SSEO_2020_ALLdata %>% summarise(distinct_dives = n_distinct(Dive))

#This output was sent to Ivy for species distribution model
write.csv(SSEO_2020_ALLdata,"outputs/SPECIES_SSEO_2020_divenum_corrected.csv")
```

## Data Issues

- The transect number and dive number were flipped for
  `SL_2020_SSEO_Dive_31_09-34-49.000.avi` — corrected to dive 31, transect
  18.

## Known Problems

None identified beyond the dive 31 fix above.

## Data Standardization

`SSEO_2020` already matches the same column structure as `EYKT_2023`, so it
doesn't need the full column-comparison check used in earlier years — only
the columns that were added to `EYKT_2023` along the way need to be added
here.

```{r standardize-2020}
SSEO_2020 <- SSEO_2020 %>% 
  mutate(mgmt_area = "SSEO",
       time_mins = NA,
       period_time_mins = NA,
       rms_2_mm = NA,
       time_akdt = NA)
```

# 2022 — CSEO & NSEO

## Raw Data

### CSEO

`File path: M:\ROVSurvey\2022\CSEO 2022\R Data Files\SPECIES_CSEO_2022_summary`

The times in the original Excel files were never converted to HHMMSS, so
when combined in R originally, the times were still incorrect. I was unable
to convert the time in the combined file, so I converted the time in the
original species review `.csv` files and recombined them using the code
`2022_CSEO_MergingCSVFilesCode_SpeciesReviewData`.

```{r import-2022-cseo}
CSEO_2022_ALLdata <- read.csv("outputs/SPECIES_CSEO_2022.csv") %>% 
  # need to fix the dive and transect numbers for dive 17 - renamed from 17b to 17
  mutate(Dive = case_when(
    Filename == "SL_2022_CSEO_Dive_17b_07-57-18.000.avi" ~ 17,TRUE ~ as.numeric(Dive))) %>%  
  #dive number and transect number were both NA - changed to what is on the dive log
  mutate(Dive = case_when(
    Filename == "SL_2022_CSEO_Dive_18_14-14-08.000.avi" ~ 18,TRUE ~ as.numeric(Dive))) %>%  
  mutate(Transect.Number = case_when(
    Filename == "SL_2022_CSEO_Dive_18_14-14-08.000.avi" ~ 15,TRUE ~ as.numeric(Transect.Number))) %>% 
  mutate(Transect.Number = case_when(
    Filename == "SL_2022_CSEO_Dive_22_15-14-56.000.avi" ~ 23,TRUE ~ as.numeric(Transect.Number)))

unique(CSEO_2022_ALLdata$Dive)
CSEO_2022_ALLdata %>% summarise(distinct_dives = n_distinct(Dive))
CSEO_2022_ALLdata %>% summarise(distinct_trans = n_distinct(Transect.Number))

#This output was sent to Ivy for species distribution model
write.csv(CSEO_2022_ALLdata,"outputs/SPECIES_CSEO_2022_summary_time_corrected.csv")
```

### NSEO

`File path: M:\ROVSurvey\2022\NSEO 2022\SPECIES REVIEW\SPECIES_NSEO_2022_summary`

```{r import-2022-nseo}
NSEO_2022_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2022_NSEO/species_NSEO_2022.csv")
```

## Data Issues

- **CSEO:** timestamps were re-converted to HHMMSS in the original per-dive
  `.csv` files (not the combined file), then recombined.
- **CSEO:** dive renamed from 17b to 17 for
  `SL_2022_CSEO_Dive_17b_07-57-18.000.avi`.
- **CSEO:** dive and transect number were both `NA` for
  `SL_2022_CSEO_Dive_18_14-14-08.000.avi` — filled in from the dive log as
  dive 18, transect 15.
- **CSEO:** transect number corrected to 23 for
  `SL_2022_CSEO_Dive_22_15-14-56.000.avi`.

## Known Problems

None identified beyond the fixes above.

## Data Standardization

`CSEO_2022` and `NSEO_2022` already match the same column structure as
`EYKT_2023`, so they don't need the full column-comparison check used in
earlier years — only the columns that were added to `EYKT_2023` along the
way need to be added here.

```{r standardize-2022}
# CSEO 2022 
CSEO_2022 <- CSEO_2022 %>% 
  mutate(mgmt_area = "CSEO",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         time_akdt = NA)

unique(CSEO_2022$dive)

# NSEO 2022 
NSEO_2022 <- NSEO_2022 %>% 
  mutate(mgmt_area = "NSEO",
         time_mins = NA,
         period_time_mins = NA,
         rms_2_mm = NA,
         time_akdt = NA)
```

# 2023 — EYKT

## Raw Data

`File path: M:\ROVSurvey\2023\EYKT\Final files for Analysis\SPECIES_EYKT_2023_summary`

```{r import-2023}
EYKT_2023_ALLdata <- read.csv("data/SPECIES_REVIEW_DATA/2023_EYKT/species_EYKT_2023.csv")
```

## Data Issues

None identified in code for this year.

## Known Problems

None identified for this year.

## Data Standardization

`EYKT_2023` serves as the template for every other year in this document
(see the "Overview" and each year's standardization section above). Its own
extra columns (`mgmt_area`, `time_mins`, `period_time_mins`, `rms_2_mm`,
`time_akdt`) were added back in the **2012** standardization step, since
that's where it was first compared against another year.

# Combine All Years

## Assemble the List and Clean Names

```{r combine-list}
all_dfs <- list(
  CSEO_2012 = CSEO_2012_ALLdata,
  SSEO_2013 = SSEO_2013_ALLdata,
  EYKT_2015 = EYKT_2015_ALLdata,
  CSEO_2016 = CSEO_2016_ALLdata,
  NSEO_2016 = NSEO_2016_ALLdata,
  EYKT_2017 = EYKT_2017_ALLdata,
  NSEO_2018 = NSEO_2018_ALLdata,
  CSEO_2018 = CSEO_2018_ALLdata,
  SSEO_2018 = SSEO_2018_ALLdata,
  EYKT_2019 = EYKT_2019_ALLdata,
  SSEO_2020 = SSEO_2020_ALLdata,
  CSEO_2022 = CSEO_2022_ALLdata,
  NSEO_2022 = NSEO_2022_ALLdata, 
  EYKT_2023 = EYKT_2023_ALLdata) 
```

Now we need to combine the dataframes into one massive dataframe. This is
going to require a lot of data manipulation.

```{r clean-names}
all_dfs <- lapply(all_dfs, clean_names)

# Here I am going to remove the columns that start with X or "user"- these are 
# empty columns and not needed
all_dfs <- lapply(all_dfs, function(df) {
  df[!grepl("^(x|user)", names(df), ignore.case = TRUE)]})

#Unpacks each cleaned dataframe back into your global environment
list2env(all_dfs, envir = .GlobalEnv)

lapply(all_dfs, names)
```

## Recombine Standardized Dataframes

```{r recombine-edited}
#recombine them in this list
all_dfs_edited <- list(
  CSEO_2012,
  SSEO_2013,
  EYKT_2015,
  CSEO_2016,
  NSEO_2016,
  EYKT_2017,
  NSEO_2018,
  CSEO_2018,
  SSEO_2018,
  EYKT_2019,
  SSEO_2020,
  CSEO_2022,
  NSEO_2022, 
  EYKT_2023) 
```

## Final Standardization

`EYKT_2023` is used as the final template — all dataframes are aligned to
its column set and coerced to character to prevent type conflicts before
binding.

```{r final-standardization}
# Set EYKT_2023 as the template
template_cols <- names(EYKT_2023)

# Clean and align all data frames to match template
all_dfs_aligned <- lapply(all_dfs_edited, function(df) {
  
  # Add any missing columns
  missing_cols <- setdiff(template_cols, names(df))
  df[missing_cols] <- NA
  
  # Reorder columns to match template
  df <- df[, template_cols, drop = FALSE]
  
  # Convert all columns to character to prevent type conflicts
  df[] <- lapply(df, as.character)
  
  return(df)
})

# COMBINE THE DATA
ROV_species_review_all_years <- bind_rows(all_dfs_aligned)
```

# Cleaning the Data

```{r clean-data}
ROV_species_review_all_years <- ROV_species_review_all_years %>% 
  mutate(year = as.character(year),
       dive_no = as.character(dive),
       transect_no = as.character(transect_number),
       species = as.character(species),
       code = as.character(code),
       length_mm = as.numeric(as.character(length_mm)),
       precision_mm = as.numeric(as.character(precision_mm)),
       rms_mm = as.numeric(as.character(rms_mm)),
       vert_dir_deg = as.character(vert_dir_deg),
       vert_dir_deg = na_if(vert_dir_deg, NA), 
       vert_dir_deg = as.numeric(as.character(vert_dir_deg))) %>% 
  mutate(tape_reader = case_when(
    tape_reader %in% c("Kristen", "Kristen Green ") ~ "Kristen Green",
    tape_reader %in% c("Jennifer Stahl", "Jenny") ~ "Jenny Stahl",
    tape_reader == "asia" ~ "Asia",
    tape_reader == "LauraColeman" ~ "Laura Coleman",
    TRUE ~ tape_reader),
    activity = case_when(
      activity == "fish seeking cover" ~ "Fish seeking cover",
      activity == "Fish milling" ~ "Fish milling/hovering",
      activity == "resting on bottom" ~ "Fish resting on bottom",
      activity == "Chase other" ~ "Fish chasing other fish",
      activity %in% c("fish moving slowly into frame", "moving slowly into frame") ~ "Fish moving slowly into frame",
      activity %in% c("fish moving quickly into frame", "moving quickly into frame") ~ "Fish moving quickly into frame",
      activity %in% c("actively swimming within frame", "Fish actively swimming within frame") ~ "Fish actively swimming in frame",
      activity %in% c("fish moving quickly out of frame", "Fish moving quickly out of frame.") ~ "Fish moving quickly out of frame",
      activity %in% c("milling", "milling/hovering") ~ "Fish milling/hovering",
      TRUE ~ activity),
    dive_type = case_when(
      dive_type == "Grouundtruth" ~ "Groundtruth",
      TRUE ~ dive_type)) %>% 
  filter(!dive_type %in% c("Groundtruth", "Exploratory")) %>% 
  mutate(dive_trans = paste(dive,transect_number, sep = "_"))

write.csv(ROV_species_review_all_years,"outputs/ROV_species_review_all_years.csv")
```

# Data Exploration

```{r explore-uniques}
names(ROV_species_review_all_years)

unique(ROV_species_review_all_years$year)
unique(ROV_species_review_all_years$mgmt_area)
unique(ROV_species_review_all_years$tape_reader)
unique(ROV_species_review_all_years$family)
unique(ROV_species_review_all_years$genus)
unique(ROV_species_review_all_years$species)
unique(ROV_species_review_all_years$code)
unique(ROV_species_review_all_years$stage)
unique(ROV_species_review_all_years$activity)
unique(ROV_species_review_all_years$dive_type)
```

```{r explore-summary-tables}
transects_per_area_per_year <- ROV_species_review_all_years %>%
  group_by(mgmt_area,year) %>%
  filter(!is.na(year)) %>% 
  summarise(num_dives = n_distinct(dive_trans)) %>%
  arrange(mgmt_area)

summary_table <- ROV_species_review_all_years %>%
  filter(!is.na(year), species == 145) %>%
  group_by(mgmt_area, year) %>%
  summarise(
    n_ad_su_yelloweye = sum(stage %in% c("AD", "SU")),
    n_juv_yelloweye   = sum(stage == "JV"),
    .groups = "drop"
  ) %>%
  arrange(mgmt_area, year)

summary_table <- summary_table %>%
  left_join(transects_per_area_per_year, by = c("mgmt_area", "year"))
```

The number of yelloweye does not match the number reported in the SAFE
reports. This could be a variety of things. In earlier years with the ROV,
fish were excluded based on size, but the raw files need to be checked.

# Activity Analysis

Since the adoption of the ROV in 2012, an average of 78% of all yelloweye
rockfish from the surveys moved minimally or slowly. Of those slow-moving
specimens, approximately 70% did not display directional movements (i.e.,
they were milling or resting on the bottom).

The above text is from the draft of the ROP. I am assuming Kelli or Phil
completed it, but I don't know where this analysis was done. Below I
replicated this analysis and updated the statement above in the ROP, which
should be published in 2026.

```{r activity-overall}
activity <- ROV_species_review_all_years %>%
  filter(genus=="yelloweye",
         !activity %in% c(""," "),
         !is.na(activity)) %>%  
  mutate(activity_new = case_when(
    activity %in% c("Fish milling/hovering",
                    "Fish resting on bottom",
                    "Fish moving slowly into frame",
                    "Fish moving slowly out of frame",
                    "Fish milling")~ "Not Moving or Minimal Movement",
    activity %in% c("Fish actively swimming in frame",
                    "Fish chasing other fish",
                    "Fish moving quickly into frame",
                    "Fish seeking cover",
                    "Fish being chased",
                    "Fish moving quickly out of frame",
                    "Fish actively swimming within frame",
                    "Feeding",
                    "Passing") ~ "Moving",
    activity == "Attracted" ~ "Attracted",
    TRUE ~ activity)) %>% 
  group_by(activity_new) %>% 
  summarize(count = n()) %>% 
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(percentage))

# Overall activity distribution
ggplot(activity, aes(x = reorder(activity_new, -percentage), y = percentage, fill = activity_new)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 4) +
  labs(title = "Yelloweye Rockfish Activity Distribution (All Years)",
    x = "Activity Category",
    y = "Percentage of Observations",
    fill = "Activity") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

Here I am asking the question — what is the percentage breakdown of the
inactive fish? What activity are they doing?

> **Note:** the code below has been adjusted from the original version.
> `group_by(inactive)` alone dropped the `activity_new` column before the
> plot tried to use it, which crashed the knit
> (`object 'activity_new' not found`). Grouping by both `inactive` and
> `activity_new` keeps the detail the plot needs, while `percentage` is
> still computed as a share of the overall `Moving`/`No Movement` totals.
> Flagging this in case the intent was actually to restrict this chart to
> *only* the "Not Moving" rows and show their relative behavior split —
> let me know if you'd rather have that version instead.

```{r activity-inactive}
inactive <- ROV_species_review_all_years %>%
  filter(genus == "yelloweye",
         !activity %in% c("", " "),
         !is.na(activity)) %>%
  mutate(
    activity_new = case_when(
      activity %in% c(
        "Fish milling/hovering",
        "Fish resting on bottom",
        "Fish moving slowly into frame",
        "Fish moving slowly out of frame",
        "Fish milling") ~ "Not Moving or Minimal Movement",
      activity %in% c(
        "Fish actively swimming in frame",
        "Fish chasing other fish",
        "Fish moving quickly into frame",
        "Fish seeking cover",
        "Fish being chased",
        "Fish moving quickly out of frame",
        "Fish actively swimming within frame",
        "Feeding",
        "Passing") ~ "Moving",
      activity == "Attracted" ~ "Attracted", TRUE ~ activity),
    inactive = case_when(activity %in% c(
        "Fish milling/hovering",
        "Fish resting on bottom",
        "Fish milling") ~ "No Movement",TRUE ~ "Moving")) %>%
  group_by(inactive, activity_new) %>% 
  summarize(count = n(), .groups = "drop") %>% 
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(percentage))

# Breakdown of minimal movement behaviors
ggplot(inactive, aes(x = reorder(activity_new, -percentage), y = percentage, fill = activity_new)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 4) +
  labs(title = "Behavior Breakdown: Not Moving or Minimal Movement",
    x = "Specific Inactive Behaviors",
    y = "Percentage within Minimal Movement Group",
    fill = "Behavior") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  coord_flip()  # Flip for readability
```
